
# COURSERA : Reproducible research - Peer Assessment 1

> Loading and preprocess the data

Ensure the required files are in source directory
read the source files

```{r}
	activity <- read.csv("activity.csv",stringsAsFactors = FALSE)
```

Summary of read source file - "activity.csv"

```{r}
str(activity)
head(activity)
```


## Questions section 1

> 1.1 Make a histogram of the total number of steps taken each day

Calculate Total number of steps taken per day, using aggregate summation 

```{r}
dailysteps <- aggregate(steps~date,activity,sum)
str(dailysteps)
head(dailysteps)
```

Create a histogram

```{r}
hist(dailysteps$steps, main = "Histogram of total number of steps taken - Per day", xlab="Total number of steps")
```

> 1.2 Calculate and report the mean and median total number of steps taken per day

```{r}
summary(dailysteps$steps)
mean(dailysteps$steps)
median(dailysteps$steps)
```

## Questions section 2


> 2.1 What is the average daily activity pattern?

To get average daily activity, we sum the steps per interval for all days. But, Using 'Function=mean'

```{r}
intervalsteps <- aggregate(steps~interval,activity,FUN=mean)

plot(intervalsteps$interval,intervalsteps$steps,type="l",xlab="Daily Intervals",ylab="Average steps per interval")
```

> 2.2 which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

We Calculate the interval using the "max()" function.
by Subsisting the data frame to where the 'steps' value in data is at its max.

```{r}
Max.interval<-subset(intervalsteps,steps==max(intervalsteps$steps))
print(Max.interval)
```

##  Questions section 3


> 3.1 Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum(is.na(activity$steps))  

```

> 3.2 Devise a strategy for filling in all of the missing values in the dataset. 

The 'activity' data frame before processing
```{r}
summary(activity)
head(activity)
```

Strategy used is to replace occurring NA with the averaged value for that interval across all dates.

This is achieved by going through all rows using a FOR loop.
Then, for all rows where there is 'NA' for steps,
we replace it with the value of steps from 'intevalsteps' data calculated before,
<B>making sure that it is for the relevant 'Time interval' 
by finding a matching value of the row 1 value of 'interval' and the interval from 'activity' of corresponding NA.</B>

```{r}
for (i in 1:nrow(activity) )  #Loop 'i' : Go from row 1 to end of Data.frame
  {
		if( is.na(activity[i,1]) )  # Check the value in (Row 'i', Column 1) of activity data frame, TRUE ?
			{
				activity[i,1] <- intervalsteps [
									   which(activity[i,3]==intervalsteps$interval),2
									  ]
			# replace the NA, with the value found in 
      #intervalsteps[where row 1 value is same as interval of the activty of the correponding NA, col 2 ]
			}
 	}
```   

> 3.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.

The 'activity' data frame after processing

```{r}  
summary(activity)
head(activity)
```


> 3.4a Make a histogram of the total number of steps taken each day

We calculate the total number of steps by using the aggregate function for 'steps VS date', using 'summation' function.

```{r}
dailysteps.no.NA <- aggregate(steps~date,activity,sum)
head(dailysteps.no.NA)
```

Creating the histogram.

```{r}
hist(dailysteps.no.NA$steps,main = "Histogram of total number of steps taken - Per day, (NA's substituted)", xlab="Total number of steps")
```

> 3.4b Calculate and report the mean and median total number of steps taken per day.

```{r}
summary(dailysteps.no.NA$steps)
mean(dailysteps.no.NA$steps)
median(dailysteps.no.NA$steps)
```

> 3.4c Do these values differ from the estimates from the first part of the assignment? 


```{r}
par(mfrow=c(2,1),mar=c(4,4,2,2),oma=c(1,1,1,1))
hist(dailysteps$steps,main="Daily steps from raw database",xlab="Intervals of the day")
hist(dailysteps.no.NA$steps,main="Daily steps with NA's replaced",,xlab="Intervals of the day")
```

```{r}
summary(dailysteps$steps)
summary(dailysteps.no.NA$steps)

```

*<B>What is the impact of imputing missing data on the estimates of the total daily number of steps?*</B>

Impact is : Though mean is same, the median is shifted.


##  Questions section 4


> 4.1 Are there differences in activity patterns between weekdays and weekends?

we first convert the 'date' in activity data frame from 'Character' to 'Date' data type.
Then add a new column called "Day" in data frame which is filled with the relevant day against the date
using the Weekdays() function.

```{r}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
activity$day <- weekdays(activity$date)
str(activity)
```


Since the day values are "Mon, Tue, Wed.. " etc. we write code to separate it into two data frame.
'Weekdays' and 'Weekends'

```{r}
activity.weekdays <- activity[
					                      c( grep("Monday|Tuesday|Wednesday|Thursday|Friday",activity$day) ) 
                                #if any row has any of these its weekday
				                  	, ]
activity.weekends <- activity[
				                      	c( grep("Sunday|Saturday",activity$day) )          
                                #if any row has any of these its a weekend
				                  	, ]
nrow(activity)
str(activity.weekends)
str(activity.weekdays)
```

*We can check accuracy by adding the row vaules of two subsets and comparing to orginial*


To calculate averages, the data again needs to be averaged against interval to get the trend.
So we repeat aggregation on the subsets.


```{r}

weekday.averages <- aggregate(steps~interval,activity.weekdays,FUN=mean)
weekend.averages <- aggregate(steps~interval,activity.weekends,FUN=mean)

str(weekday.averages)
str(weekend.averages)
```

Plot the data

```{r}
plot(weekday.averages$interval,weekday.averages$steps,type="l",xlab="Daily Intervals",ylab="Average Steps",main="Weekday (Black), Weekend (Blue)")

lines(weekend.averages$interval,weekend.averages$steps,type="l",col="blue")
```

<B>Result:</B>
We can see that mid day activity during weekend is higher.

*Maybe because the person doesn't have to sit in the office on weekends !

But, the he highest intensity within a certain interval is still in the weekday.

*Maybe he has to walk stairs up to his office on 200the floor ??


#THE END#
